Salmon straying description

### 2017-05-02 manuscript/msv1.tex
heritability fix
mtheta

Here we consider a local population $N_i$ and a remote population $N_j$, each with a trait value $x_i$ and $x_j$ that determines the recruitment rate, are assumed to be normally distributed with means $\mu_i$ and $\mu_j$ and a standard deviation $\sigma$, and are subject to selection across ecological timescales.
Each population in its respective site has an optimum trait value $\theta_i$ where the rate of recruitment is maximized if $x_i = \theta_i$.
If there is no straying between these populations (such that they are independent), then the trait value should evolve towards the optimum ($x_i \rightarrow \theta_i$) and the recruitment rate will become maximized.
If there is straying between populations at rate $m$, then the traits in each respective location will be pulled away from the optimum, and recruitment rates will be lower, i.e. have lower fitness.
As $m$ becomes large, the populations should act as a single population.

Using the discrete population framework from Shelton and Mangel (2011), we have the following population dynamic equation for a local population $i$ and the population that is straying into it $j$.

\begin{align}
  N_i(t+1) &= \left(N_i(t) - m N_i(t) + m N_j(t) \right){\rm e}^{-Z} \\ \nonumber
  &+ \left((R_i[\mu_i(t)|\theta_i]+\tilde{P}) (N_i(t) - m N_i(t))+ (R_i[\mu_j(t)|\theta_i]+\tilde{P}) m N_j(t)\right) {\rm e}^{-\beta (N_i(t) - m N_i(t)+ m N_j(t))}
\end{align}

\noindent where the first line determines the number of individuals that survive from time $t$ to time $t+1$, and the second and third line accounts for the recruitment of individuals from the surviving population from time $t$ to time $t+1$.
Moreover, there is a demographic process error term included in the above dynamic equation, where $\tilde{P}\sim \mathcal{N}(0,s)$ where in the below example $s=0.001$

Survival: the probability of survival is ${\rm e}^{-Z}$... if $Z=0.5$ (Shelton and Mangel 2011), the probability of survival is 60\%. 
This survival probability is enacted on 1) The number of local individuals that return minus the number local individuals that stray ($N_i(t) - mN_i(t)$), in addition to the number of remote individuals that immigrate into the local patch $mN_j(t)$, and it is assumed that they all have equal probability of surviving.

Recruitment: The second line of the equation determines the recruitment from local individuals $N_i(t)-mN_i(t)$ and incoming strays $mN_j(t)$, as a function of their mean trait value at time $t$ and given the local trait optimum $\theta_i$, where the recruitment rate at low population densities is
\begin{align}
  R_i[\mu_i(t)|\theta_i] &= \int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i(t)-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i(t)|\mu_i,\sigma^2) {\rm d}x_i(t) \\ \nonumber
  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-\mu_i(t))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\}.
\end{align}
This means that it is the mismatch between the trait mean $\mu_i(t)$ and the local optimum $\theta_i$ that determines the  growth rate at low population densities.
% The compensatory effects are then determined by the exponential, following the Ricker stock-recruitment relationship.

The third line of the equation determines recruitment from straying individuals (with density $mN_j(t)$), such that the recruitment rate is determined by the mismatch between the optimum trait value $\theta_i$ and the mean trait value of the incoming individuals $\mu_j$.

As shown above, the mean traits of both populations (the local mean trait value $\mu_i$ as well as the remote mean trait value $\mu_j$) also change through time.
Because individuals from the local population are being mixed with individuals from the remote population, the resulting trait distribution is a mixed normal.
We assume that the distribution that results after the remote and local populations are mixed together \emph{and reproduce} is normal with a mean value determined from a mixed-normal distribution and constant variance over time.
This assumption may be tested a few different ways if we deem it necessary down the road - as a first order approximation, it's probably fine.
The mean trait value thus changes through time according to the difference equation

\begin{equation}
  \mu_i(t+1) = w_i\mu_i(t) + (1-w_i)\mu_j(t) + \frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)|\theta_i] + (1-w_i)R_i[\mu_j(t)|\theta_i]  \right)
  \label{eq:mu}
\end{equation}

\noindent where

\begin{equation}
  w_i=\frac{N_i(t)-m N_i(t)}{N_i(t)-m N_i(t) + m N_j(t)}.
\end{equation}

The first two factors in equation \ref{eq:mu} determine the mixed normal average of the now-mixed local and remote populations.
This mixed normal is weighted by the proportion of the population that is local and remote, respectively, which depends on the stray rate $m$.
The third part of the above equation determines how the mean trait changes through time due to natural selection (from Lande), which is proportional to the change in mean fitness with respect to the mean trait (phenotype).

You can see from the elements of the difference equations that demographics directly impact the evolution of trait means due to how the local and remote trait averages at time $t$ contribute to the trait mean at time $t+1$.

%Density dependent m

@@

% Constant m
%% Fold bifurcation @ intermediate values of m
%% PE > 1 for low values of m
%% Spike in PE at the fold bifurcation (spike in indvidual population SD)
%% Increased environmental heterogeneity leads to spikes in PE at lower stray rates
%% Effects of heritability vs. trait variation and stable states

@@

The unexpected result are the alternative stable states that arise for lower values of $m$. Just where these alternative stable states arise depends on the variance as well as the difference in optimal trait values between sites ($\theta_1-\theta_2$), but you can see that this has big impacts on the portfolio effect.
Which trajectory goes to the higher or lower alternative stable state is random - an effect of the process noise injected into the system.
The change in the PE appears to be caused by large spikes in the variance of individual population trajectories where the alternative stable states begin (at the fold bifurcation).
This is a well-known phenomenon that forms the basis for early warning signals of phase transitions.

\begin{figure}[h]
\centering
\includegraphics[width=0.4\textwidth]{fig_Density.pdf}
\caption{}
\end{figure}

The differences in optimal trait values between sites ($\theta_1-\theta_2$) influence at what straying rate the alternative stable states appear.
Large differences in $\theta_1$ and $\theta_2$ can be interpreted as high environmental heterogeneity.
A cursory analysis shows that increasing environmental heterogeneity tends to lower the stray rate value $m^*$ at which the maximum PE is expected to occur.
Moreover, if the trait variance is much larger than the difference in optimal trait values between sites (such that selection is weaker), the alternative stable state phenomenon disappears, as does the spike in the portfolio effect.

\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth]{fig_HSDcomb.png}
\caption{}
\end{figure}

>>>>

Here we consider two populations $N_1$ and $N_2$ that are separated in space, each with trait values $x_1$ and $x_2$ determining recruitment rates.
We assume that there is an optimum trait value $\theta_1$ and $\theta_2$ associated with each habitat, where recruitment is maximized if the trait value of the local population $x = \theta$.
Moreover, we assume that $x_{1,2}$ are normally distributed with means $\mu_1$ and $\mu_2$ and have a shared standard deviation $\sigma$.
As such, the recruitment rate for both populations is determined by the mean trait value of the local population, such that $r_1 = R_1[\mu_1(t)|\theta_1]$.
Trait means for each population are subject to selection, the strength of which depends on the difference between the population mean and the local trait optimum at a given point in time.

The two populations are assumed to reproduce in spatially separate sites that are close enough such that a proportion of the population $m$ can stray into the wrong site, and where mortality occurs before individuals return to spawn.
If there is no straying between these populations (such that they are independent), then the mean trait evolves towards the optimal value such that $x_1 \rightarrow \theta_1$, and the recruitment rate for that population will be maximized.
If there is straying between populations at rate $m$, then the traits in each respective location will be pulled away from the optimum, and recruitment rates will be lowered.
As $m \rightarrow 0.5$, the populations are perfectly mixed, acting as a single population.

We use the discrete Ricker population dynamic framework described by Shelton and Mangel \cite{} as the basis for our two-site model, with the added effect of the local population $N_i$ mixing with a set proportion $m$ of a remote population $N_j$ that is straying into it.
We first assume that the proportion ${\rm e}^{-Z}$ of both populations survive, and that the aggregated mix of the populations (local individuals in addition to the straying individuals) are subject to the same compensatory effects, determined by the parameter $\beta$.
For a local site $i\in(1,2)$ that collects straying individuals from a remote site $j\in(1,2)$, if $N_i$ is the local site and $N_j$ is the remote site, the difference equation that determine changes in population size is

\begin{align}
  N_i(t+1) &= \left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z} \\ \nonumber
  &+ \left((R_i[\mu_i(t)|\theta_i]+\tilde{P}) (1-m)N_i(t) + (R_i[\mu_j(t)|\theta_i]+\tilde{P}) m N_j(t)\right) {\rm e}^{-\beta ((1-m)N_i(t) + m N_j(t))},
\end{align}

\noindent where we include demographic process error with $\tilde{P}\sim \mathcal{N}(0,s)$, where $s$ is assumed to be small, and where the difference equation for $N_j$ mirrors that for $N_i$.

The combined recruitment of local individuals $(1-m)N_i(t)$ and incoming strays $mN_j(t)$, as a function of their mean trait value at time $t$ and given the local trait optimum $\theta_i$, is then

\begin{align}
  R_i[\mu_i(t)|\theta_i] &= \int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i(t)-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i(t)|\mu_i,\sigma^2) {\rm d}x_i(t) \\ \nonumber
  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-\mu_i(t))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\}.
\end{align}

\noindent As stated previously, it is the mismatch between the local trait mean $\mu_i(t)$ and the local optimum $\theta_i$ that determines the recruitment rate for the population.
The parameter $\tau$ controls the sensitivity of recruitment to changes in the mean trait value away from the optimum, which we set as $\tau=1$ here and throughout.
% The compensatory effects are then determined by the exponential, following the Ricker stock-recruitment relationship.

Because individuals from the local population are mixed with individuals from the remote population via staying, the resulting trait distribution is a mixed normal with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and for the straying individuals, $1-w_i$, where $w_i=(1-m)N_i(t)/((1-m) N_i(t) + m N_j(t))$.
We make two simplifying assumptions.
First, we assume that the distribution resulting from the mix of remote and local individuals, following reproduction, is also normal with a mean value being that of the mixed-normal.
Second, we assume that changes in trait variance through time are minimal, such that $\sigma$ is assumed to be constant.

An increasing flow of incoming strays is thus expected to pull the mean trait value of the local population away from its optimum, which will decrease its rate of recruitment.
The mean trait value thus changes through time according to the difference equation

\begin{equation}
  \mu_i(t+1) = w_i\mu_i(t) + (1-w_i)\mu_j(t) + \frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)|\theta_i] + (1-w_i)R_i[\mu_j(t)|\theta_i]  \right),
  \label{eq:mu}
\end{equation}

\noindent where the first two factors determine the mixed normal average of the now-mixed local and remote populations.
This mixed normal is weighted by the proportion of the population that is local and remote, respectively, which depends on the stray rate $m$.
The partial derivative in the Eq. \ref{eq:mu} determines how the mean trait changes through time due to natural selection (REF), which is proportional to the change in mean fitness with respect to $\mu_i$.

%Density dependent m
We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is good evidence that at least in some species the stray rate is density dependent.
Specifically, the rate at which individuals stray has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tends to decrease the rate at which individuals stray, thus reducing the overall proportion of a population that strays.
According to REF, given the probability that an individual strays $m_0$, the proportion of the local population $N_i(t)$ that strays is

\begin{equation}
  m(t) = m_0 - m_0\left(\frac{N_i(t)}{C+N_i(t)}\right)
\end{equation}

\noindent where $C$ is the half-saturation value of $N_i$ where the density-dependent stray rate decreases sharply.
We note that at the limit $C\rightarrow \infty$, the density dependent stray rate becomes constant such that $m(t) \rightarrow m_0$, and this corresponds to the original formulation where $m=m_0$.
A similar observation shows that when the population density is very high, $m(t) \rightarrow 0$, and when it is close to extinction, $m(t) \rightarrow m_0$.
Thus, for realistic population densities, $m(t) < m_0$.

%Linking m/m0 and thetadiff

@@

%The effects of trait evolution :: trait heritability, trait variance, and habitat heterogeneity
%% Bifurcation to alternative stable state occurs at higher stray rates with increased trait heritability
%% There is a peak in the portfolio effect directly before this bifurcation is crossed (early warning signal)
%% When trait heritability is high, there is a greater sensitivity of total biomass to increasing stray rates, though a LOWER sensitivity of the difference between alternative steady state values for increasing stray rates (and vice versa)
%% For low heritability (h = 0.01 - 0.2 or so), moderate and insensitive changes in total biomass, but large differences between the values of alternative steady states leads to a local maximum for PE at low-moderate stray rates
%% Increased habitat heterogeneity
  %%% 1) exaggerates lowering of mean biomass with increasing stray rates
  %%% 2) increases the likelihood of alternative stable sites & increases the differences in alternative stable state values with increasing stray rates
  %%% 3) decreases the PE (!!)

%% PE >> 1 for low values of m
%% Spike in PE at the fold bifurcation (spike in indvidual population SD)
%% Increased environmental heterogeneity leads to spikes in PE at lower stray rates
%% Effects of heritability vs. trait variation and stable states

@@

The unexpected result are the alternative stable states that arise for lower values of $m$. Just where these alternative stable states arise depends on the variance as well as the difference in optimal trait values between sites ($\theta_i-\theta_j$), but you can see that this has big impacts on the portfolio effect.
Which trajectory goes to the higher or lower alternative stable state is random - an effect of the process noise injected into the system.
The change in the PE appears to be caused by large spikes in the variance of individual population trajectories where the alternative stable states begin (at the fold bifurcation).
This is a well-known phenomenon that forms the basis for early warning signals of phase transitions.

\begin{figure}[h]
\centering
\includegraphics[width=0.4\textwidth]{figs/fig_Density.pdf}
\caption{}
\end{figure}

The differences in optimal trait values between sites ($\theta_i-\theta_j$) influence at what straying rate the alternative stable states appear.
Large differences in $\theta_i$ and $\theta_j$ can be interpreted as high environmental heterogeneity.
A cursory analysis shows that increasing environmental heterogeneity tends to lower the stray rate value $m^*$ at which the maximum PE is expected to occur.
Moreover, if the trait variance is much larger than the difference in optimal trait values between sites (such that selection is weaker), the alternative stable state phenomenon disappears, as does the spike in the portfolio effect.

\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth]{figs/fig_HSDcomb.png}
\caption{}
\end{figure}
